---
title: "Model 2: Feature Engineering for Churn Prediction"
author: "Group 6"
date: "April 14, 2025"
format: html
---

This document presents a binary classification model to predict customer churn using feature engineering. We'll create exactly 5 engineered features that maximize the Kappa score.

## Setup

Load the required libraries:

```{r setup}
library(tidyverse)
library(caret)
```

## Load Data

Load the BankChurners dataset:

```{r load-data}
# Read the data
bank <- readRDS(gzcon(url("https://cd-public.github.io/D505/dat/BankChurners.rds")))

bank <- bank %>% mutate(Revenue = Total_Revolving_Bal/5 +  Total_Trans_Ct/4 - 15 * Contacts_Count_12_mon)

# Check the data structure
glimpse(bank)
```

## Feature Engineering

Create the 5 engineered features based on extensive analysis of feature importance and correlation:

```{r feature-engineering}
bank_engineered <- bank %>%
  mutate(
    # Feature 1: Enhanced Churn Risk Score v2
    # Combines inactive months, contact frequency, relationship depth, and transaction volatility
    enhanced_churn_risk_v2 = (Months_Inactive_12_mon / 12) * (Contacts_Count_12_mon / 4) *
                          (1 / (Total_Relationship_Count + 0.1)) *
                          (1 / (abs(Total_Ct_Chng_Q4_Q1) + 0.5)) * 0.6 +
                          (Contacts_Count_12_mon / (pmax(Total_Trans_Ct, 1)) * 10 / 10) * 0.4,
    
    # Feature 2: Customer Activity Score
    # Focuses on transaction patterns and inactivity
    customer_activity_score = (log1p(Total_Trans_Ct) * 0.5) + 
                            ((Total_Ct_Chng_Q4_Q1 + 1) * 0.3) - 
                            (Months_Inactive_12_mon / 12 * 0.2),
    
    # Feature 3: Contact-to-Transaction Ratio
    # Measures support needs relative to transaction activity
    contact_transaction_ratio = Contacts_Count_12_mon / 
                              (pmax(Total_Trans_Ct, 1)) * 10,
    
    # Feature 4: Complex Engagement Score
    # Non-linear combination of key activity metrics
    complex_engagement = (Total_Trans_Ct / 70)^0.8 * 
                       (Total_Relationship_Count / 4)^0.7 * 
                       (1 - (Months_Inactive_12_mon / 12))^1.2,
    
    # Feature 5: Hybrid Score v2
    # Comprehensive engagement metric with revenue component
    hybrid_score_v2 = ((Total_Relationship_Count * 0.25) + 
                     (Total_Trans_Ct * 0.35) - 
                     (Months_Inactive_12_mon * 0.25) +
                     (Revenue / 500 * 0.15) +
                     ((Total_Trans_Ct / 70) * 
                     (Total_Relationship_Count / 4) * 
                     (12 - Months_Inactive_12_mon) / 12)) / 2
  )
```

## Select Features for Model

Extract only the engineered features and the target variable:

```{r select-features}
# Select only the engineered features plus Churn column for the model
bank_model <- bank_engineered %>%
  select(enhanced_churn_risk_v2, customer_activity_score, 
         contact_transaction_ratio, complex_engagement, 
         hybrid_score_v2, Churn)

# Preview the engineered features
head(bank_model)
```

## Explore Feature Distributions

Visualize the distributions of the engineered features by churn status:

```{r feature-visualization}
# Convert Churn to a factor for better visualization
bank_model$Churn <- factor(bank_model$Churn, levels = c("no", "yes"))

# Create a long format version for easier plotting
bank_model_long <- bank_model %>%
  pivot_longer(cols = -Churn, 
               names_to = "Feature", 
               values_to = "Value")

# Plot distributions
ggplot(bank_model_long, aes(x = Value, fill = Churn)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ Feature, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Engineered Features by Churn Status",
       x = "Feature Value", 
       y = "Density") +
  scale_fill_brewer(palette = "Set1")
```

## Train Model

Train a logistic regression model using 5-fold cross-validation:

```{r train-model}
# Set seed for reproducibility
set.seed(123)

# Train the model
model <- train(Churn ~ ., 
               data = bank_model,
               trControl = trainControl(method = "cv", number = 5),
               method = "glm",
               family = "binomial",
               maxit = 5)

# Display results
print(model)
print(paste("Kappa value:", model$results[["Kappa"]]))
```

## Feature Importance

Examine the coefficients to understand feature importance:

```{r feature-importance}
# Extract coefficients
coef_values <- coef(model$finalModel)
coef_df <- data.frame(
  Feature = names(coef_values),
  Coefficient = as.numeric(coef_values)
) %>%
  slice(-1)  # Remove intercept

# Plot feature importance
ggplot(coef_df, aes(x = reorder(Feature, abs(Coefficient)), y = Coefficient)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Feature Importance (Absolute Coefficient Values)",
       x = "Feature",
       y = "Coefficient")
```

## Save the Dataframe

Save the engineered dataset for submission:

```{r save-dataframe}
# Save the dataframe
write_rds(bank_model, file = "model_2.rds")
print("Saved dataframe to model_2.rds")
```

## Assessment Verification

Verify that our assessment matches the expected format from the assignment:

```{r assessment}
# Verify the assessment matches the expected format
assessment <- train(Churn ~ ., 
                   data = bank_model %>% select(1, 2, 3, 4, 5, grep("Churn", colnames(bank_model))),
                   trControl = trainControl(method = "cv", number = 5),
                   method = "glm",
                   family = "binomial",
                   maxit = 5)

print(assessment)
print(paste("Assessment Kappa value:", assessment$results[["Kappa"]]))
```

## Feature Descriptions

### Enhanced Churn Risk v2

This feature combines multiple risk signals: inactive months, contact frequency, relationship depth, and transaction volatility. Higher values indicate greater churn probability. The feature captures both direct risk indicators (inactivity) and their interactions.

### Customer Activity Score

This metric measures overall transaction engagement and activity trajectory. It uses log-transformed transaction count to capture diminishing returns, rewards transaction growth, and penalizes inactivity.

### Contact-to-Transaction Ratio

This ratio measures the "maintenance cost" per transaction and is a good predictor of customer dissatisfaction or account problems. Higher values often indicate customers requiring more support relative to their activity level.

### Complex Engagement

This non-linear engagement measure uses power transformations to capture the interaction between transaction activity, relationship depth, and recency. The non-linear transformation helps capture diminishing returns and interaction effects.

### Hybrid Score v2

This comprehensive engagement metric incorporates relationship count (product breadth), transaction count (activity level), inactivity penalty, revenue component (customer value), and interaction terms between transactions, relationships, and activity.
